{ \fontfamily{ptm}\selectfont Catalyst Meta Acceleration Framework: The history and the gist of it }

### 2024-11-12 Notebook LaTEX/Catalyst Accelerations/main.tex
>w<, changes and new results

\title{
    {
        \fontfamily{ptm}\selectfont 
        Catalyst Meta Acceleration Framework: The Gist of its Theories
    }
    }

@@

\noindent{\bfseries 2010 Mathematics Subject Classification:}
Primary 65K10, 90c25, 90C30; Secondary 65Y20. 
\noindent{\bfseries Keywords: } Nesterov acceleration, Proximal point method.

@@

    One would be tempting to conclude that this has sealed the ceiling for research on the topic of extending Nesterov's acceleration. 
    That is indeed correct, but not from a practical point of view. 
    Let $F: \RR^n \mapsto \RR$ be our objective function, $\mathcal J_\lambda := (I + \lambda \partial F)^{-1}$ and $\mathcal M^{\lambda}(x; y):= F(x) + \frac{1}{2\lambda}\Vert x - y\Vert^2$ then the inexact proximal point considers with error $\epsilon_k$ has the following characterizations of inexactness as put forward by Guler \cite{guler_new_1992}: 
    \begin{align*}
        & \tilde x \approx \mathcal J_\lambda y
        \\
        & 
        \dist(\partial\mathcal{M}^{\lambda}(\tilde x; y))
        \le \frac{\epsilon}{\lambda}
    \end{align*}
    However, this is troublesome because if we need to approximate the resolvent operator $\mathcal J_\lambda$, then it's probably difficult to compute the subgradient $\partial\mathcal M(\cdot; y)$, which make it difficult to know when we achieved the required exactness for a PPM evaluation. 
    Otherwise, if we already know the subgradient well, then why approximate it in the first place?

@@

    \subsection{Contributions}
        The writing is expository and won't contain major results. 
        We reviewed the literatures and faithfully reproduced some claims, in addition we give insights into understanding the claim in relations to other papers and foundational ideas in optimization.

@@

@@

\section{Lin 2015}
    This section introduced the setup of Nesterov's estimating sequence in Lin 2015 \cite{lin_universal_2015}. 
    Right at the beginning we warn the readers about the following: 
    \begin{enumerate}
        \item The proofs in HongZhou Lin's original paper of Universal Catalyst is depressingly long and complicated. This is a result of using the constructive approach of Nesterov's estimating sequence. 
        \item It's context specific for controlling the errors of inexact proximal point evaluations are being controlled. He hinted at the way of controlling and tracking the errors from inexact proximal point evaluations, but it's context specific. He only illustrated the use of the meta acceleration on their own method called: ``Proximal MISO", but in general the problem still remains open. 
        \item We will provide proofs to clarify some of their proofs and compare with existing proofs and drawing references in the literatures in the appendix. 
    \end{enumerate}
    Let's assume $F$ is a $\mu \ge 0$ strongly convex function. 
    Throughout this section we make the following notations
    \begin{align*}
        \mathcal M^{\kappa^{-1}}(x; y) &:= F(x) + \frac{\kappa}{2}\Vert x - y\Vert^2, 
        \\
        \mathcal J_{\kappa^{-1}} y &:= \argmin_x \mathcal M^{\kappa^{-1}} (x, y). 
    \end{align*}
    Their algorithm is almost exactly the same as Nesterov's 2.2.20 \cite{nesterov_lectures_2018} which we stated in the definition below: 
    \begin{definition}[Lin's accelerated proximal point method]
        Let the initial estimate be $x_0 \in \RR^n$, fix parameters $\kappa$ and $\alpha_0$. 
        Let $(\epsilon_k)_{k \ge 0}$ be an error sequence chosen for the evaluation for inexact proximal point method. 
        Initialize $x_0 = y_0$, then the algorithm generates $(x_k, y_k)$ satisfies for all $k \ge 1$
        \begin{align*}
            & \text{find } x_k \approx \mathcal J_{\kappa^{-1}} y_{k - 1} \text{ such that } \mathcal M^{\kappa^{-1}}(x_k, y_{k - 1}) - \mathcal M^{\kappa^{-1}}(\mathcal J_{\kappa^{-1}}y_{k - 1}, y_{k - 1}) \le \epsilon_k
            \\
            & \text{find } \alpha_k \in (0, 1) \text{ such that } \alpha_k^2 = (1 - \alpha_k)\alpha_{k - 1}^2 + (\mu/(\mu + \kappa)) 
            \\
            & 
            y_{k} = x_k + \frac{\alpha_{k - 1}(1 - \alpha_{k - 1})}{\alpha_{k - 1}^2 + \alpha_k}(x_k - x_{k - 1}). 
        \end{align*}
    \end{definition}
    \begin{remark}
        The most exciting thing about this algorithm is the similarity it has compares to 
        \hyperref[def:acc-prox-grad-alg]{Definition \ref*{def:acc-prox-grad-alg}}. 
        The only difference here is the inclusion of an inexact proximal point step, and the Lipschitz constant $L$ is absent instead it has $\kappa + \mu$.

        Lin 2015 \cite{lin_universal_2015} commented on the second last paragraph on page 4, and here we quote:
        \par
        ``The choice of the sequence $(\epsilon_k)_{k \ge 0}$ is also subjected to discussion since the quantity $F(x_0) - F^*$ is unknown beforehand. Nevertheless, an upper bound may be used instead, which will only affects the corresponding constant in (7). Such an upper bounds can typically be obtained by computing a duality gap at $x_0$, or by using additional knowledge about the objective. For instance, when $F$ is non-negative, we may simply choose $\epsilon_k = (2/9)F(x_0)(1 - \rho)^k$".

        This comment has upmost practical importance because it tells us how to bound the error $\epsilon_k$ to achieve accelerated convergence rate. 
        However, it's still elusive for general in practical applications. 
        To see how the error is controlled for the inexact proximal point evaluation, we refer the readers to Lemma B.1 in Lin's 2015 paper \cite{lin_universal_2015}. 
    \end{remark}
    For theoretical interests, there is a major difference between Lin's approach and Guler's approach. 
    Lin didn't formulate any of the intractable quantities in the definitions for his Nesterov's estimating sequence $\phi_k$. 
    One major innovations here is Lemma A.7 in Lin's 2015 paper \cite{lin_universal_2015}.

\section{Non-convex Extension of Catalyst Acceleration}
    The non-convex extension of Catalyst acceleration by Lin 2018\cite{lin_catalyst_2018} remains similar to the algorithm in the convex case in his 2015 paper \cite{lin_universal_2015}. 
    The algorithm is now adapted to handle function with unknown weak convexity constant $\rho$ using the process called Auto Adapt subroutine. 
    There is no accelerated rate for the non-convex case since it's impossible, but the theoretical convergence to stationary point is claimed for the non-convex case.

>>>>

\title{
    {
        \fontfamily{ptm}\selectfont 
        Catalyst Meta Acceleration Framework: The history and the gist of it
    }
    }

@@

% \noindent{\bfseries 2010 Mathematics Subject Classification:}
% Primary 65K10, 90c25, 90C30; Secondary 65Y20. 
% \noindent{\bfseries Keywords: } Nesterov acceleration, Proximal point method.

@@

    One would be tempting to conclude that this has sealed the ceiling for research on the topic of extending Nesterov's acceleration. 
    That is indeed correct, but not from a practical point of view. 
    Let $F: \RR^n \mapsto \RR$ be our objective function, $\mathcal J_\lambda := (I + \lambda \partial F)^{-1}$ and $\mathcal M^{\lambda}(x; y):= F(x) + \frac{1}{2\lambda}\Vert x - y\Vert^2$ then the inexact proximal point considers with error $\epsilon_k$ has the following characterizations of inexactness as put forward by Guler \cite{guler_new_1992}: 
    \begin{align*}
        & \tilde x \approx \mathcal J_\lambda y
        \\
        & 
        \dist(\partial\mathcal{M}^{\lambda}(\tilde x; y))
        \le \frac{\epsilon}{\lambda}. 
    \end{align*}
    However, this is troublesome because if we need to approximate the resolvent operator $\mathcal J_\lambda$, then it's probably difficult to compute the subgradient $\partial\mathcal M(\cdot; y)$, which make it difficult to know when we achieved the required exactness for a PPM evaluation. 
    Otherwise, if we already know the subgradient well, then why approximate it in the first place?

@@

    \subsection{Contributions}
        The writing is expository and comprehensive. 
        We reviewed the literatures and faithfully reproduced some claims, in addition we give insights into understanding the claim in relations to other papers and foundational ideas in optimization. 
        Three papers by Guler \cite{guler_new_1992} and Lin \cite{lin_universal_2015} and Paquette et al. \cite{lin_catalyst_2018} together with Nesterov's \cite{nesterov_lectures_2018} method of estimating sequence which is introduced in his book will be covered in detail for this report.

        We only cover innovations in the theoretical aspect of Catalyst Acceleration. 
        Detailed applications and specific examples will be out of the scope of this paper simply because there will be too many and requires a comprehensive understanding on a different collection of papers.

@@

    In Guler's work, these intractable quantities were built into the Nesterov's estimating sequence which means it impossible to control how $\epsilon_k \rightarrow 0$. 

@@

\section{Lin 2015}
    This section introduced the setup of Nesterov's estimating sequence in Lin 2015 \cite{lin_universal_2015}. 
    Right at the beginning we warn the readers about the following: 
    \begin{enumerate}
        \item The proofs in HongZhou Lin's original paper of Universal Catalyst is depressingly long and complicated. This is a result of using the constructive approach of Nesterov's estimating sequence. 
        \item It's context specific for controlling the errors of inexact proximal point evaluations are being controlled. He hinted at the way of controlling and tracking the errors from inexact proximal point evaluations, but it's context specific. He only illustrated the use of the meta acceleration on their own method called: ``Proximal MISO", but in general the problem still remains open. 
        \item We will provide proofs to clarify some of their proofs and compare with existing proofs and drawing references in the literatures in the appendix. 
    \end{enumerate}
    Let's assume $F$ is a $\mu \ge 0$ strongly convex function. 
    Throughout this section we make the following notations
    \begin{align*}
        \mathcal M^{\kappa^{-1}}(x; y) &:= F(x) + \frac{\kappa}{2}\Vert x - y\Vert^2, 
        \\
        \mathcal J_{\kappa^{-1}} y &:= \argmin_x \mathcal M^{\kappa^{-1}} (x, y). 
    \end{align*}
    Their algorithm is almost exactly the same as Nesterov's 2.2.20 \cite{nesterov_lectures_2018} which we stated in the definition below: 
    \begin{definition}[Lin's accelerated proximal point method]
        Let the initial estimate be $x_0 \in \RR^n$, fix parameters $\kappa$ and $\alpha_0$. 
        Let $(\epsilon_k)_{k \ge 0}$ be an error sequence chosen for the evaluation for inexact proximal point method. 
        Initialize $x_0 = y_0$, then the algorithm generates $(x_k, y_k)$ satisfies for all $k \ge 1$
        \begin{align*}
            & \text{find } x_k \approx \mathcal J_{\kappa^{-1}} y_{k - 1} \text{ such that } \mathcal M^{\kappa^{-1}}(x_k, y_{k - 1}) - \mathcal M^{\kappa^{-1}}(\mathcal J_{\kappa^{-1}}y_{k - 1}, y_{k - 1}) \le \epsilon_k
            \\
            & \text{find } \alpha_k \in (0, 1) \text{ such that } \alpha_k^2 = (1 - \alpha_k)\alpha_{k - 1}^2 + (\mu/(\mu + \kappa)) 
            \\
            & 
            y_{k} = x_k + \frac{\alpha_{k - 1}(1 - \alpha_{k - 1})}{\alpha_{k - 1}^2 + \alpha_k}(x_k - x_{k - 1}). 
        \end{align*}
    \end{definition}
    \begin{remark}
        The most exciting thing about this algorithm is the similarity it has compares to 
        \hyperref[def:acc-prox-grad-alg]{Definition \ref*{def:acc-prox-grad-alg}}. 
        The only difference here is the inclusion of an inexact proximal point step, and the Lipschitz constant $L$ is absent instead it has $\kappa + \mu$. 
        Evaluating $x_k \approx \mathcal J_{\kappa^{-1}}y_{k - 1}$ is also possible because the function $\mathcal M^{\kappa^{-1}}(\cdot, y_{k - 1})$ is strongly convex, hence its optimality gap can be bounded via trackable quantity $\partial \mathcal M^{\kappa^{-1}}(x_k, y_{k - 1})$.

        Controlling the error sequence $\epsilon_k$ however is a whole new business. 
        Lin 2015 \cite{lin_universal_2015} commented on the second last paragraph on page 4, and here we quote:
        \par
        ``The choice of the sequence $(\epsilon_k)_{k \ge 0}$ is also subjected to discussion since the quantity $F(x_0) - F^*$ is unknown beforehand. Nevertheless, an upper bound may be used instead, which will only affects the corresponding constant in (7). Such an upper bounds can typically be obtained by computing a duality gap at $x_0$, or by using additional knowledge about the objective. For instance, when $F$ is non-negative, we may simply choose $\epsilon_k = (2/9)F(x_0)(1 - \rho)^k$".

        This comment has upmost practical importance because it tells us how to bound the error $\epsilon_k$ to achieve accelerated convergence rate. 
        In theory, $\epsilon_k$ decreases at a rate related to $F(x_0) - F^*$. 
        It requires some knowledge about $F^*$ in prior. 
        Therefore, controlling $\epsilon_k$ is still elusive in general in a practical context. 
        To see how the error is controlled for the inexact proximal point evaluation, we refer the readers to Lemma B.1 in Lin's 2015 paper \cite{lin_universal_2015}. 
    \end{remark}
    For theoretical interests, there is a major difference between Lin's approach and Guler's approach. 
    Lin didn't formulate any of the intractable quantities in the definitions for his Nesterov's estimating sequence $\phi_k$. 
    One major innovation is Lemma A.7 in Lin's 2015 paper \cite{lin_universal_2015}. 
    The lemma allows the analysis Nesterov's estimating sequence to be carried through without using intractable quantities: $\mathcal M^{\kappa^{-1}}(\mathcal J_{k^{-1}}y_{k - 1}, y_{k - 1}), \mathcal J_{\kappa^{-1}}y_{k - 1}$.

\section{Non-convex Extension of Catalyst Acceleration}
    The non-convex extension of Catalyst acceleration by Lin 2018 \cite{lin_catalyst_2018} remains similar to the algorithm in the convex case in his 2015 paper \cite{lin_universal_2015}. 
    The algorithm is now adapted to handle function with unknown weak convexity constant $\rho$ using the process called Auto Adapt subroutine. 
    There is no accelerated rate for the non-convex case since it's impossible, but the theoretical convergence to stationary point is claimed for the non-convex case.
